Exploring probability distributions for bivariate temporal granularities

### 2019-12-15 main.Rmd
measured variable section update

## Summarizing properties of the distribution

Estimating the distribution of the measured variable can give valuable indication of features like central tendency, skewness, multimodality, tail behavior or variation. The measured variable in the considered data structure can be dicrete or continuous. The probability distribution of them could be studied by estimating the probability density function (continuous) or probability mass function (dicrete). The approach can be parametric or non-parametric. A non-parametric approach makes less rigid assumptions about the distribution of the observed data making the data speak more for itself. These resonate more with the notion of EDA which advocates exploring data for patterns and relationships without requiring prior hypotheses. The entire distribution could be estimated using different density estimation methods or summarized using empirical quantiles. Several forms of  non-parametric density estimators and quantile estimators exist in the literature, the most commonly used form ( @Silverman1986-cl,  @Jones1992-pj) are as follows:

@@

Hence, the ingredients for the plots are subsets of numbers (or categorical measurements), and the choice of faceting or axis alignment for the subset strata. The basic plot choice is display of density. Different plots might be appropriate depending on which features of the density we are interested to look at, the levels of time granularities being plotted, how granularities interact and number of observations available. We will look at each of these aspects separately and analyze the effect of each on visualization. While we accomplish that, we need to bear in mind that any customized species of visualization might be constructed by varying the set of mappings between data properties and visual attributes such as position, shape and color. We are trying to address problems where even with good aesthetic qualities and good data, the graph might be confusing or misleading because of how people process the information.

@@

## Choice of

@@

There are several ways to visualize statistical distributions. The entire distribution can be displayed through density estimates using non-paramteric density estimation methods or only summarized values can be displayed using descriptive statistics. The latter do not allow us to see the shape, skewness, nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter, where some specific probabilities representing typical and extreme behavior are on focus. Non-parametric density based plots can show the entire distribution, unlike discretizing and showing only parts of it. These displays depend on the specific estimation procedure (e.g., kernel density estimate) as well as on additional smoothing parameters and hence can vary for the same data set. Furthermore, these methods are subject to the same sample size restrictions and challenges that apply to any density estimation. In practice, the densities are estimated reasonably with at least 30 observations. However, even with large sample size, choosing too large a value for bandwidth can cause the estimate to oversmooth the data. The visualization tools based on descriptive statistics display only actual observations from the sample, and remain free of any tuning parameters or Kernels.

@@

<!-- The probabilities displayed are based on actual data, which is aligned to principles that governed Tukey’s original boxplot. They can be useful in spotting multimodality, however, they might become obscuring with too many categories and information on the entire distribution can add to cognitive load. Also, the probabilities in density plot are estimated through kernel density estimation and thus makes assumptions when selecting kernel or bandwidth. As a result, the plot shows smooth summaries based on the assumptions and not only on actual observations. -->

@@

Given a context, it is good to be conversant with the benefits and challenges while choosing a distribution plot. As a general rule, if we have too many categories, the quantile plots are useful for comparing patterns, whereas, other more involved methods of plotting are useful for studying anomalies, outlier or multimodal behavior.

>>>>

## Measured variable

<!-- Moreover, the measured variable in the considered data structure can be dicrete or continuous. -->

Summarizing properties of the distribution of the measured variable can give valuable indication of features like central tendency, skewness, multimodality, tail behavior or variation. With huge amount of data being available, mean, median or any one summary statistic might not be enough to understand the different properties of the measured variable. Instead the distribution could be summarized using some summary statistics together, each of which can highlight different properties of the data. Tukey's five number summary, letter values or quantiles can serve as examples of distributional summaries. Another way of summarizing could be estimating the probability density function (continuous) or probability mass function (dicrete). The approach can be parametric or non-parametric. A non-parametric approach makes less rigid assumptions about the distribution of the observed data making the data speak more for itself. These resonate more with the notion of EDA which advocates exploring data for patterns and relationships without requiring prior hypotheses. Several forms of  non-parametric density estimators and quantile estimators exist in the literature, the most commonly used form ( @Silverman1986-cl,  @Jones1992-pj) are as follows:

@@

<!-- Hence, the ingredients for the plots are subsets of numbers (or categorical measurements), and the choice of faceting or axis alignment for the subset strata. The basic plot choice is display of density. Different plots might be appropriate depending on which features of the density we are interested to look at, the levels of time granularities being plotted, how granularities interact and number of observations available. We will look at each of these aspects separately and analyze the effect of each on visualization. While we accomplish that, we need to bear in mind that any customized species of visualization might be constructed by varying the set of mappings between data properties and visual attributes such as position, shape and color. We are trying to address problems where even with good aesthetic qualities and good data, the graph might be confusing or misleading because of how people process the information. -->

@@

## Choice of statistical transformations and geometric objects

@@

There are several ways to summarize a statistical distributions. The entire distribution can be estimated through density estimates using non-paramteric density estimation methods or only summarized using descriptive statistics. The latter do not allow us to see the shape, skewness, nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter, where some specific probabilities representing typical and extreme behavior are on focus. Non-parametric density based plots can show the entire distribution, unlike discretizing and showing only parts of it. These displays depend on the specific estimation procedure (e.g., kernel density estimate) as well as on additional smoothing parameters and hence can vary for the same data set. In practice, the densities are estimated reasonably with at least 30 observations. However, even with large sample size, choosing too large a value for bandwidth can cause the estimate to oversmooth the data. The summaries based on descriptive statistics remain free of any tuning parameters or Kernels.

@@

<!--  Furthermore, these methods are subject to the same sample size restrictions and challenges that apply to any density estimation.  -->
<!-- The probabilities displayed are based on actual data, which is aligned to principles that governed Tukey’s original boxplot. They can be useful in spotting multimodality, however, they might become obscuring with too many categories and information on the entire distribution can add to cognitive load. Also, the probabilities in density plot are estimated through kernel density estimation and thus makes assumptions when selecting kernel or bandwidth. As a result, the plot shows smooth summaries based on the assumptions and not only on actual observations. -->

@@

Each type of density display has different parameters. These are quantities that need to be estimated given the data. Each is equipped with some benefits and challenges. For example, if we have too many categories in the variables mapped to facet or x-axis, the quantile plots might be useful for comparing patterns, whereas, other more involved methods of plotting are useful for studying anomalies, outlier or multimodal behavior.
